---
title: "Final Report: Diabetes: Association Between Glycemic Control, Lipid Profile, and Kidney Function Across Diabetes Classifications"
date: 2026-06-04
author: "Sherif Hamidu"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: false
    theme: cosmo

editor: visual
---

# Final Report: Diabetes: Association Between Glycemic Control, Lipid Profile, and Kidney Function Across Diabetes Classifications

## Introduction

Diabetes mellitus is a chronic metabolic disorder characterized by persistent hyperglycemia that can lead to both macrovascular and microvascular complications. Glycated hemoglobin (HbA1c) is the standard clinical marker used to assess long-term glycemic control and is associated with the risk of diabetes-related complications. Poor glycemic control has been linked to abnormalities in lipid metabolism, including elevated cholesterol, triglycerides, and low-density lipoprotein (LDL), as well as reduced high-density lipoprotein (HDL). These metabolic disturbances contribute to cardiovascular disease, the leading cause of morbidity among individuals with diabetes.

Diabetes also affects kidney function, with elevated serum urea and creatinine serving as indicators of renal impairment. Understanding how glycemic control relates to lipid abnormalities and kidney function across different diabetes classifications may help identify patients at greater risk of complications.

This study investigates the association between HbA1c, lipid profile, and kidney function across diabetes classifications.

### Research Question

How are HbA1c levels associated with lipid profile and kidney function across different diabetes classifications?

### Hypothesis

Individuals with poorer glycemic control (higher HbA1c) will have a more adverse lipid profile and poorer kidney function, and these relationships will differ among diabetes classifications.

## Methods

### Data Source

The dataset consists of clinical measurements obtained from individuals representing different diabetes classifications. Variables include glycemic control (HbA1c), lipid profile biomarkers (total cholesterol, triglycerides, HDL, LDL, and VLDL), and kidney function biomarkers (urea and creatinine). These variables were analyzed to investigate the relationship between glycemic control, lipid metabolism, and kidney function.

###Software

All analyses were performed in R using Quarto for reproducible reporting.

Load Required Packages

The following R packages were used for data manipulation, visualization, and statistical analysis.

```{r}
#| message: false
#| warning: false
# Load a library
# Install packages
#install.packages("tidyverse")
#install.packages("ggpubr")
#install.packages("corrplot")
#install.packages("psych")
#install.packages("GGally")
#| message: false
#| warning: false

# Load packages
library(tidyverse)
library(ggpubr)
library(corrplot)
library(psych)
library(GGally)
library(ggplot2)
```

\### Import the Dataset

The diabetes dataset was imported into R and inspected to verify that it was read correctly.

```{r}
getwd()

list.files()

Diabetes <- read.csv("Dataset of Diabetes .csv")
```

### Data Inspection

The structure of the dataset was explored to understand the available variables, their data types, and the overall dimensions of the dataset.

```{r}
glimpse(Diabetes)

head(Diabetes)

str(Diabetes)

summary(Diabetes)

dim(Diabetes)
```

### Missing Value Assessment

Before performing statistical analyses, the dataset was examined for missing values.

```{r}
colSums(is.na(Diabetes))
```

A dataset with minimal or no missing values improves the reliability of statistical analyses and visualizations.

### Variables Included in this Study

## Study Variables

The analysis focused on biomarkers representing three major biological domains:

| Biological Domain | Variables                |
|-------------------|--------------------------|
| Glycemic Control  | HbA1c                    |
| Lipid Metabolism  | Chol, TG, HDL, LDL, VLDL |
| Kidney Function   | Urea, Cr                 |

### Data Preparation

To simplify subsequent analyses, only the variables relevant to the research question were retained.

```{r}
Diabetes_selected <-
Diabetes %>%
select(
CLASS,
HbA1c,
Chol,
TG,
HDL,
LDL,
VLDL,
Urea,
Cr
)

head(Diabetes_selected)

### Data Normalization

###Because the biomarkers are measured on different clinical scales, a normalized dataset was created using Z-score standardization. This preprocessing step allows variables to be compared on the same scale while preserving the original dataset for statistical analyses.

Diabetes_normalized <-
Diabetes_selected %>%
mutate(
across(
HbA1c:Cr,
scale
)
)

head(Diabetes_normalized)


```

The normalized dataset was created to facilitate comparison across biomarkers. However, all descriptive statistics, visualizations, and statistical analyses presented in this report were performed using the original clinical measurements to preserve clinical interpretability.

## Results

This section presents the findings of the analyses performed to investigate the relationship between glycemic control (HbA1c), lipid profile, and kidney function across diabetes classifications. The results are organized according to the study's research questions.

```{r}
---
## Descriptive Statistics

#Before conducting statistical analyses, summary statistics were calculated to describe the distribution of HbA1c, lipid profile markers, and kidney function markers included in this study.

# Descriptive statistics for continuous variables

psych::describe(
  Diabetes_selected[, c("HbA1c", "Chol", "TG", "HDL", "LDL", "VLDL", "Urea", "Cr")]
)

```

These descriptive statistics demonstrate the variation present among glycemic, lipid, and renal biomarkers. Because these variables are measured using different clinical units, normalization was performed separately to allow comparison across biomarkers.

------------------------------------------------------------------------

# Research Question 1

## Does glycemic control (HbA1c) differ across diabetes classifications?

To answer this question, the distribution of HbA1c was first examined using a histogram. HbA1c values were then compared among diabetes classifications using boxplots. Finally, a one-way Analysis of Variance (ANOVA) was performed to determine whether the observed differences were statistically significant.

------------------------------------------------------------------------

### Distribution of HbA1c

The histogram below illustrates the overall distribution of HbA1c values within the study population. Examining the distribution helps determine whether HbA1c is approximately normally distributed or exhibits skewness.

```{r}
ggplot(Diabetes_selected,
       aes(x = HbA1c)) +
  
  geom_histogram(
    aes(y = after_stat(density)),
    bins = 20,
    fill = "steelblue",
    color = "black",
    alpha = 0.8
  ) +
  
  geom_density(
    color = "red",
    linewidth = 1.2
  ) +
  
  theme_minimal(base_size = 14) +
  
  labs(
    title = "Distribution of HbA1c",
    x = "HbA1c (%)",
    y = "Density"
  )
```

**Interpretation**

The histogram demonstrates the distribution of HbA1c values across all participants. HbA1c values span a broad range, indicating substantial variability in glycemic control within the study population. The distribution suggests the presence of individuals with both normal glucose regulation and poor glycemic control, supporting further investigation of metabolic differences associated with HbA1c levels.

Take-home message: This dataset contains participants with varying degrees of glycemic control, making it suitable for studying metabolic changes associated with diabetes.

------------------------------------------------------------------------

### HbA1c Across Diabetes Classifications

To compare glycemic control among diabetes classifications, HbA1c values were visualized using boxplots.

```{r}
diabetes <- Diabetes %>%
  mutate(
    CLASS = factor(
      CLASS,
      levels = c("N", "P", "Y"),
      labels = c("Non-diabetic", "Prediabetic", "Diabetic")
    )
  )

hbA1c_box_jitter <- ggplot(diabetes, aes(x = CLASS, y = HbA1c, fill = CLASS)) +
  geom_boxplot(width = 0.6, outlier.shape = NA) +
  stat_summary(
    fun = mean,
    geom = "point",
    shape = 23,
    size = 3,
    fill = "white",
    color = "black"
  ) +
  geom_jitter(
    aes(color = CLASS),
    width = 0.15,   # small horizontal spread
    alpha = 0.25,   # low opacity → less visual density
    size = 1        # small point size
  ) +
  scale_fill_brewer(palette = "Set2") +
  scale_color_brewer(palette = "Set2", guide = "none") +
  labs(
    title = "HbA1c across diabetes classifications",
    x = "Diabetes classification",
    y = "HbA1c (%)"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "none",
    plot.title = element_text(hjust = 0.5, face = "bold"),
    axis.title.x = element_text(face = "bold"),
    axis.title.y = element_text(face = "bold")
  )

hbA1c_box_jitter
```

**Interpretation**

HbA1c differed significantly among diabetes classifications. Individuals classified as diabetic exhibited substantially higher HbA1c levels than prediabetic and non-diabetic participants. The highly significant ANOVA result (p \< 0.001) confirms that glycemic control varies across diabetes classifications and supports the use of HbA1c as a distinguishing clinical marker.

Take-home message: As expected, diabetes is characterized by chronically elevated HbA1c, providing the foundation for investigating downstream metabolic alterations.

------------------------------------------------------------------------

### Statistical Comparison of HbA1c Across Diabetes Classifications

A one-way Analysis of Variance (ANOVA) was performed to determine whether the mean HbA1c differed significantly among diabetes classifications.

```{r}
anova <- aov(
  HbA1c ~ CLASS,
  data = Diabetes
)

summary(anova)
```

**Interpretation**

HbA1c differed significantly across non-diabetic, prediabetic, and diabetic groups, confirming that the clinical classification reflects progressively worse glycemic control.

# Research Question 2

## Is HbA1c associated with lipid profile?

### Correlation Between HbA1c and Lipid Biomarkers

A correlation matrix was generated to evaluate the strength and direction of relationships between HbA1c, lipid markers, and kidney function biomarkers.

```{r}
# Select variables for correlation analysis

library(corrplot)

corrdata <- Diabetes %>%
  select(HbA1c, Chol, TG, HDL, LDL, VLDL, Urea, Cr)



# Calculate correlation matrix
cormatrix <- cor(corrdata, use = "complete.obs")


# Visualize correlations

corrplot(
  cormatrix,
  method = "color",
  type = "upper",
  addCoef.col = "black",
  number.cex = 0.7,
  tl.col = "black",
  tl.srt = 45,
  tl.cex = 0.9,
  col = colorRampPalette(c("steelblue4", "white", "tomato3"))(200),
  diag = FALSE
)

```

**Figure 3. Correlation matrix showing relationships between glycemic control, lipid profile, and kidney function biomarkers.**

Correlation matrix illustrating relationships among HbA1c, lipid biomarkers, and kidney function markers. Positive correlations indicate biomarkers that increase together, whereas negative correlations indicate inverse relationships. HbA1c showed stronger positive associations with triglycerides and total cholesterol than with kidney biomarkers, suggesting that lipid metabolism is more closely linked to worsening glycemic control.



------------------------------------------------------------------------

### Relationship Between HbA1c and Individual Lipid Markers

To further examine specific relationships identified in the correlation analysis, scatterplots were generated between HbA1c and each lipid biomarker. Linear regression lines were added to visualize the direction of association.

------------------------------------------------------------------------


```{r}
### Class Label
library(dplyr)
library(forcats)

Diabetes <- Diabetes %>%
  mutate(
    Class_Label = recode_factor(
      CLASS,
      N = "Non-diabetic",
      P = "Prediabetic",
      Y = "Diabetic"
    )
  )
```

## HbA1c and Total Cholesterol

```{r}
#| message: false
#| warning: false
chol_plot <- ggplot(Diabetes, aes(x = HbA1c, y = Chol, color = Class_Label)) +
  geom_point(alpha = 0.30, size = 1.5) +
  geom_smooth(method = "lm", se = TRUE, linewidth = 1) +
  scale_color_brewer(palette = "Set2") +
  theme_minimal(base_size = 12) +
  labs(
    title = "HbA1c vs Total Cholesterol",
    x = "HbA1c (%)",
    y = "Total chol",
    color = "Diabetes class"
  )
chol_plot
```

------------------------------------------------------------------------

## HbA1c and Triglycerides

```{r}
#| message: false
#| warning: false
tg_plot <- ggplot(Diabetes, aes(x = HbA1c, y = TG, color = Class_Label)) +
  geom_point(alpha = 0.30, size = 1.5) +
  geom_smooth(method = "lm", se = TRUE, linewidth = 1) +
  scale_color_brewer(palette = "Set2") +
  theme_minimal(base_size = 12) +
  labs(
    title = "HbA1c vs Total Triglycerides",
    x = "HbA1c (%)",
    y = "Total TG",
    color = "Diabetes class"
  )
tg_plot
```

------------------------------------------------------------------------

## HbA1c and HDL

```{r}
#| message: false
#| warning: false
hdl_plot <- ggplot(Diabetes, aes(x = HbA1c, y = HDL, color = Class_Label)) +
  geom_point(alpha = 0.30, size = 1.5) +
  geom_smooth(method = "lm", se = TRUE, linewidth = 1) +
  scale_color_brewer(palette = "Set2") +
  theme_minimal(base_size = 12) +
  labs(
    title = "HbA1c vs Total HDL",
    x = "HbA1c (%)",
    y = "Total HDL",
    color = "Diabetes class"
  )
hdl_plot
```

------------------------------------------------------------------------

## HbA1c and LDL

```{r}
#| message: false
#| warning: false
ldl_plot <- ggplot(Diabetes, aes(x = HbA1c, y = LDL, color = Class_Label)) +
  geom_point(alpha = 0.30, size = 1.5) +
  geom_smooth(method = "lm", se = TRUE, linewidth = 1) +
  scale_color_brewer(palette = "Set2") +
  theme_minimal(base_size = 12) +
  labs(
    title = "HbA1c vs Total LDL",
    x = "HbA1c (%)",
    y = "Total LDL",
    color = "Diabetes class"
  )
ldl_plot
```

------------------------------------------------------------------------

## HbA1c and VLDL

```{r}
#| message: false
#| warning: false
vldl_plot <-ggplot(Diabetes, aes(x = HbA1c, y = VLDL, color = Class_Label)) +
  geom_point(alpha = 0.30, size = 1.5) +
  geom_smooth(method = "lm", se = TRUE, linewidth = 1) +
  scale_color_brewer(palette = "Set2") +
  theme_minimal(base_size = 12) +
  labs(
    title = "HbA1c vs Total VLDL",
    x = "HbA1c (%)",
    y = "Total VLDL",
    color = "Diabetes class"
  )
vldl_plot
```

------------------------------------------------------------------------

## Combined Lipid Profile Visualization

To improve visualization and allow direct comparison among lipid biomarkers, individual scatterplots were combined into a single multi-panel figure.

```{r}
library(ggpubr)

lipid_panel <- ggarrange(
  chol_plot,
  tg_plot,
  hdl_plot,
  ldl_plot,
  vldl_plot,
  ncol = 2,
  nrow = 3,
  labels = LETTERS[1:5]
)

lipid_panel
```

**Figure 4. Association between HbA1c and lipid profile biomarkers.**

Correlation matrix illustrating relationships among HbA1c, lipid biomarkers, and kidney function markers. Positive correlations indicate biomarkers that increase together, whereas negative correlations indicate inverse relationships. HbA1c showed stronger positive associations with triglycerides and total cholesterol than with kidney biomarkers, suggesting that lipid metabolism is more closely linked to worsening glycemic control.

Take-home message: Triglycerides showed the strongest association with HbA1c, highlighting insulin resistance as a major driver of altered lipid metabolism in diabetes.

------------------------------------------------------------------------

# Research Question 3

## Is HbA1c associated with kidney function?

Diabetic kidney disease is one of the most common complications of diabetes and is strongly influenced by prolonged exposure to elevated blood glucose levels. Chronic hyperglycemia can contribute to renal injury through oxidative stress, inflammation, and vascular dysfunction.

Serum urea and creatinine are commonly used biomarkers for evaluating kidney function. Therefore, this study examined whether HbA1c levels were associated with changes in kidney function markers.

------------------------------------------------------------------------

## Relationship Between HbA1c and Kidney Function Biomarkers

Scatterplots with linear regression lines were generated to evaluate the association between HbA1c and kidney function markers.

------------------------------------------------------------------------

## HbA1c and Urea

```{r}
#| message: false
#| warning: false
urea_plot <- ggplot(Diabetes, aes(x = HbA1c, y = Urea, color = Class_Label)) +
  geom_point(alpha = 0.30, size = 1.5) +
  geom_smooth(method = "lm", se = TRUE, linewidth = 1) +
  scale_color_brewer(palette = "Set2") +
  theme_minimal(base_size = 12) +
  labs(
    title = "HbA1c vs Total Urea",
    x = "HbA1c (%)",
    y = "Total Urea",
    color = "Diabetes class"
  )
urea_plot
```

**Figure 5A. Relationship between HbA1c and serum urea levels.**

------------------------------------------------------------------------

## HbA1c and Creatinine

```{r}
#| message: false
#| warning: false
#| paged-print: false
cr_plot <- ggplot(Diabetes, aes(x = HbA1c, y = Cr, color = Class_Label)) +
  geom_point(alpha = 0.30, size = 1.5) +
  geom_smooth(method = "lm", se = TRUE, linewidth = 1) +
  scale_color_brewer(palette = "Set2") +
  theme_minimal(base_size = 12) +
  labs(
    title = "HbA1c vs Total Cr",
    x = "HbA1c (%)",
    y = "Total Cr",
    color = "Diabetes class"
  )
cr_plot
```

**Figure 5B. Relationship between HbA1c and serum creatinine levels.**

------------------------------------------------------------------------

## Combined Kidney Function Visualization

To improve comparison between renal biomarkers, both kidney function plots were displayed together.

```{r}
ggarrange(
urea_plot,
cr_plot,
ncol = 2
)
```

**Figure 5C. Association between HbA1c and kidney function biomarkers.**
Scatterplots showing the association between HbA1c and renal biomarkers. Neither urea nor creatinine demonstrated a significant relationship with HbA1c in this cohort. Although chronic hyperglycemia is known to impair kidney function over time, these findings suggest that measurable renal dysfunction may not yet be present in many participants.

Take-home message: Metabolic dysregulation appears before detectable kidney impairment in this study population.

------------------------------------------------------------------------

# Research Question 4

## Can lipid profile and kidney function biomarkers predict HbA1c levels?

While individual correlations describe relationships between HbA1c and individual biomarkers, diabetes-related metabolic abnormalities often occur together. Therefore, a multiple linear regression model was performed to determine whether lipid profile markers and kidney function biomarkers collectively predict HbA1c levels.

The model included:

- Total cholesterol (Chol)
- Triglycerides (TG)
- HDL
- LDL
- VLDL
- Urea
- Creatinine (Cr)

HbA1c was used as the dependent variable because it represents long-term glycemic control.

------------------------------------------------------------------------

## Multiple Linear Regression Model

In this model, HbA1c is the outcome we want to predict. The predictor variables include all lipid biomarkers and the two kidney function biomarkers. The model estimates the independent contribution of each biomarker while adjusting for the effects of the others.

The following model was used:

\[ HbA1c = \beta\_0 + \beta\_1(Chol) + \beta\_2(TG) + \beta\_3(HDL) + \beta\_4(LDL) + \beta\_5(VLDL) + \beta\_6(Urea) + \beta\_7(Cr) \]

```{r}
#| message: false
#| warning: false
model <-

lm(
HbA1c ~ Chol +
TG +
HDL +
LDL +
VLDL +
Urea +
Cr,

data = Diabetes
)


library(broom)
library(ggplot2)

tidy(model,
     conf.int=TRUE) |>
  
  filter(term!="(Intercept)") |>
  
  ggplot(
    aes(
      estimate,
      reorder(term,estimate)
    )
  )+
  
  geom_point(size=3)+
  
  geom_errorbarh(
    aes(
      xmin=conf.low,
      xmax=conf.high
    ),
    height=.2
  )+
  
  geom_vline(
    xintercept=0,
    linetype=2
  )+
  
  theme_minimal()+
  
  labs(
    title="Regression Coefficients",
    x="Estimate",
    y=""
  )
```
**Figure 6 Regressional Model**
Regression coefficient plot summarizing the independent association between each biomarker and HbA1c after adjusting for the other variables in the model. Each point represents the estimated regression coefficient, while the horizontal bars represent the 95% confidence intervals. Biomarkers whose confidence intervals do not cross zero are considered statistically significant independent predictors of HbA1c.

Take-home message: Triglycerides and total cholesterol are the strongest independent biomarkers associated with worsening glycemic control, while kidney biomarkers remain relatively unchanged in this cohort.


------------------------------------------------------------------------

## Discussion

This analysis examined how glycemic control, measured by HbA1c, relates to lipid profile and kidney function across diabetes classifications. As expected, HbA1c increased from non-diabetic to prediabetic to diabetic groups, confirming its role as a robust marker of worsening glycemic status. The correlation and regression results show that HbA1c is more strongly associated with lipid abnormalities than with the kidney function markers included in this dataset.

Total cholesterol and triglycerides emerged as significant predictors of HbA1c, suggesting that poorer glycemic control tends to co-occur with a more adverse lipid profile. This is consistent with the concept of diabetic dyslipidemia, in which chronic hyperglycemia and insulin resistance drive atherogenic changes in lipid metabolism and increase cardiovascular risk. In contrast, urea and creatinine showed weaker and non-significant independent associations with HbA1c, indicating that these routine renal markers may not closely track glycemic status in this cross-sectional dataset. Kidney involvement in diabetes may require more sensitive measures or additional clinical information, such as eGFR, albuminuria, disease duration, and comorbid conditions, to be fully characterized.


## Conclusion

In summary, this study shows that HbA1c differs clearly across diabetes classifications and is more closely linked to lipid abnormalities than to the kidney function markers evaluated. Higher HbA1c is associated with an unfavorable lipid profile, particularly elevated cholesterol and triglycerides, highlighting the importance of assessing both glycemic control and lipid status in individuals with diabetes. Routine measures of urea and creatinine, however, did not demonstrate strong independent relationships with HbA1c, suggesting that renal risk may not be captured adequately by these markers alone. These findings support an integrated metabolic approach to diabetes care that combines glycemic monitoring with detailed lipid assessment, while future work should incorporate more sensitive renal indices and broader clinical covariates to clarify the relationship between glycemic control and kidney disease.


## References
1. American Diabetes Association Professional Practice Committee. Standards of Care in Diabetes—2026. Diabetes Care. 2026;49(Suppl 1):S1–S200.

2. International Diabetes Federation. IDF Diabetes Atlas. 11th ed. Brussels: International Diabetes Federation; 2025.

3. Goldberg IJ. Diabetic dyslipidemia: causes and consequences. J Clin Endocrinol Metab. 2001;86(3):897–904.

4. Chahil TJ, Ginsberg HN. Diabetic dyslipidemia. Endocrinol Metab Clin North Am. 2006;35(3):491–510.

5. Vergès B. Pathophysiology of diabetic dyslipidaemia: where are we? Diabetologia. 2015;58(5):886–899.

6. Ali S, et al. Diabetic dyslipidemia: a screening tool for cardiovascular risk. J Family Med Prim Care. 2021;10(12):4431–4436.

7. Wan EYF, et al. Diabetes with poor-control HbA1c is cardiovascular disease risk equivalent: a population-based retrospective cohort study. BMJ Open Diabetes Res Care. 2023;11(1):e003124.

8. Butalia S, et al. Association between hemoglobin A1c and development of cardiovascular disease and mortality. Diabetes Care. 2024;47(5):e60–e68.

9. Romera IC, et al. The association between HbA1c levels and the risk of myocardial infarction and stroke in type 2 diabetes. Cardiovasc Diabetol. 2025;24(1):45.

10. Dabla PK. Renal function in diabetic nephropathy. World J Diabetes. 2010;1(2):48–56.

11. McGill JB, et al. Making an impact on kidney disease in people with type 2 diabetes. BMJ Open Diabetes Res Care. 2022;10(4):e002957.

12. Kidney Disease: Improving Global Outcomes (KDIGO) CKD Work Group. KDIGO 2024 Clinical Practice Guideline for the Evaluation and Management of Chronic Kidney Disease. Kidney Int Suppl. 2024;14(1):1–115.

13. American Academy of Family Physicians. Diabetic kidney disease: diagnosis, treatment, and prevention. Am Fam Physician. 2019;99(12):751–759.

14. Tsimihodimos V, et al. Dyslipidemia in type 2 diabetes: pathophysiology and management. Curr Pharm Des. 2018;24(29):3650–3662.


**Last updated:** June 2026
