---
title: "Final Report: [Diabetes: Association Between Glycemic Control, Lipid Profile, and Kidney Function Across Diabetes Classifications]"
date: 2026-06-04
author: "Sherif Hamidu"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: false
editor: visual
---

# Final Report: \[Diabetes: Association Between Glycemic Control, Lipid Profile, and Kidney Function Across Diabetes Classifications\]

## Introduction

Diabetes mellitus is a chronic metabolic disorder characterized by persistent hyperglycemia that can lead to both macrovascular and microvascular complications. Glycated hemoglobin (HbA1c) is the standard clinical marker used to assess long-term glycemic control and is associated with the risk of diabetes-related complications. Poor glycemic control has been linked to abnormalities in lipid metabolism, including elevated cholesterol, triglycerides, and low-density lipoprotein (LDL), as well as reduced high-density lipoprotein (HDL). These metabolic disturbances contribute to cardiovascular disease, the leading cause of morbidity among individuals with diabetes.

Diabetes also affects kidney function, with elevated serum urea and creatinine serving as indicators of renal impairment. Understanding how glycemic control relates to lipid abnormalities and kidney function across different diabetes classifications may help identify patients at greater risk of complications.

This study investigates the association between HbA1c, lipid profile, and kidney function across diabetes classifications.

### Research Question

How are HbA1c levels associated with lipid profile and kidney function across different diabetes classifications?

### Hypothesis

Individuals with poorer glycemic control (higher HbA1c) will have a more adverse lipid profile and poorer kidney function, and these relationships will differ among diabetes classifications.

## Methods

### Data Source

The dataset consists of clinical measurements obtained from individuals representing different diabetes classifications. Variables include glycemic control (HbA1c), lipid profile biomarkers (total cholesterol, triglycerides, HDL, LDL, and VLDL), and kidney function biomarkers (urea and creatinine). These variables were analyzed to investigate the relationship between glycemic control, lipid metabolism, and kidney function.

###Software

All analyses were performed in R using Quarto for reproducible reporting.

Load Required Packages

The following R packages were used for data manipulation, visualization, and statistical analysis.

```{r}
# Load a library
# Install packages
#install.packages("tidyverse")
#install.packages("ggpubr")
#install.packages("corrplot")
#install.packages("psych")
#install.packages("GGally")

# Load packages
library(tidyverse)
library(ggpubr)
library(corrplot)
library(psych)
library(GGally)
library(ggplot2)
```

Import the Dataset

The diabetes dataset was imported into R and inspected to verify that it was read correctly.

```{r}
getwd()

list.files()

Diabetes <- read.csv("Dataset of Diabetes .csv")
```

### Data Inspection

The structure of the dataset was explored to understand the available variables, their data types, and the overall dimensions of the dataset.

```{r}
glimpse(Diabetes)

head(Diabetes)

str(Diabetes)

summary(Diabetes)

dim(Diabetes)
```

### Missing Value Assessment

Before performing statistical analyses, the dataset was examined for missing values.

```{r}
colSums(is.na(Diabetes))
```

A dataset with minimal or no missing values improves the reliability of statistical analyses and visualizations.

### Variables Included in this Study

| Variable | Description                  | Category          |
|----------|------------------------------|-------------------|
| CLASS    | Diabetes classification      | Grouping variable |
| HbA1c    | Glycated hemoglobin          | Glycemic control  |
| Chol     | Total cholesterol            | Lipid marker      |
| TG       | Triglycerides                | Lipid marker      |
| HDL      | High-density lipoprotein     | Lipid marker      |
| LDL      | Low-density lipoprotein      | Lipid marker      |
| VLDL     | Very low-density lipoprotein | Lipid marker      |
| Urea     | Blood urea                   | Kidney function   |
| Cr       | Serum creatinine             | Kidney function   |

### Data Preparation

To simplify subsequent analyses, only the variables relevant to the research question were retained.

```{r}
Diabetes_selected <-
Diabetes %>%
select(
CLASS,
HbA1c,
Chol,
TG,
HDL,
LDL,
VLDL,
Urea,
Cr
)

head(Diabetes_selected)

### Data Normalization

###Because the biomarkers are measured on different clinical scales, a normalized dataset was created using Z-score standardization. This preprocessing step allows variables to be compared on the same scale while preserving the original dataset for statistical analyses.

Diabetes_normalized <-
Diabetes_selected %>%
mutate(
across(
HbA1c:Cr,
scale
)
)

head(Diabetes_normalized)

###The normalized dataset was created to facilitate comparison across biomarkers. However, all descriptive statistics, visualizations, and statistical analyses presented in this report were performed using the original clinical measurements to preserve clinical interpretability.
```

## Results

This section presents the findings of the analyses performed to investigate the relationship between glycemic control (HbA1c), lipid profile, and kidney function across diabetes classifications. The results are organized according to the study's research questions.

```{r}
---
## Descriptive Statistics

#Before conducting statistical analyses, summary statistics were calculated to describe the distribution of HbA1c, lipid profile markers, and kidney function markers included in this study.

# Descriptive statistics for continuous variables

psych::describe(
  Diabetes_selected[, c("HbA1c", "Chol", "TG", "HDL", "LDL", "VLDL", "Urea", "Cr")]
)

```

The descriptive statistics provide an overview of the central tendency, variability, and range of each clinical biomarker. These summaries help identify differences in measurement scales and provide context for the subsequent analyses.

---

# Research Question 1

## Does glycemic control (HbA1c) differ across diabetes classifications?

To answer this question, the distribution of HbA1c was first examined using a histogram. HbA1c values were then compared among diabetes classifications using boxplots. Finally, a one-way Analysis of Variance (ANOVA) was performed to determine whether the observed differences were statistically significant.

---

### Distribution of HbA1c

The histogram below illustrates the overall distribution of HbA1c values within the study population. Examining the distribution helps determine whether HbA1c is approximately normally distributed or exhibits skewness.

```{r}
ggplot(Diabetes,
       aes(HbA1c)) +

  geom_histogram(
    bins = 30,
    fill = "steelblue",
    color = "black"
  ) +

  labs(
    title = "Figure 1. Distribution of HbA1c",
    x = "HbA1c (%)",
    y = "Frequency"
  ) +

  theme_classic()
```

**Interpretation**

The histogram provides an overview of the distribution of HbA1c values in the study population. The shape of the distribution can be used to assess whether HbA1c is approximately normally distributed or skewed, which is useful when interpreting subsequent statistical analyses.

---

### HbA1c Across Diabetes Classifications

To compare glycemic control among diabetes classifications, HbA1c values were visualized using boxplots.

```{r}
ggplot(Diabetes,
       aes(CLASS,
           HbA1c,
           fill = CLASS)) +

  geom_boxplot() +

  labs(
    title = "Figure 2. HbA1c Across Diabetes Classifications",
    x = "Diabetes Classification",
    y = "HbA1c (%)"
  ) +

  theme_classic() +

  theme(
    legend.position = "none"
  )
```

**Interpretation**

The boxplots display the median, interquartile range, overall spread, and potential outliers for HbA1c within each diabetes classification. Visual differences between groups suggest whether glycemic control varies among diabetes classifications.

---

### Statistical Comparison of HbA1c Across Diabetes Classifications

A one-way Analysis of Variance (ANOVA) was performed to determine whether the mean HbA1c differed significantly among diabetes classifications.

```{r}
anova <- aov(
  HbA1c ~ CLASS,
  data = Diabetes
)

summary(anova)
```

**Interpretation**

The ANOVA results indicate whether mean HbA1c differs significantly across diabetes classifications.

- If **p < 0.05**, conclude that HbA1c differs significantly among at least one of the diabetes classifications.
- If **p ≥ 0.05**, conclude that there is no statistically significant difference in HbA1c among the groups.

These findings provide the foundation for investigating whether differences in glycemic control are associated with alterations in lipid profile and kidney function.

---
# Research Question 2

## Is HbA1c associated with lipid profile?

### Poor glycemic control in diabetes is frequently accompanied by abnormalities in lipid metabolism. Elevated HbA1c has been associated with increased triglycerides, cholesterol, LDL, and VLDL levels, while HDL levels may decrease as metabolic control worsens.

### To investigate whether glycemic control is associated with lipid abnormalities, correlations between HbA1c and lipid biomarkers were examined. Individual relationships were further explored using scatterplots with linear regression lines.

---

## Correlation Between HbA1c and Lipid Biomarkers

A correlation matrix was generated to evaluate the strength and direction of relationships between HbA1c, lipid markers, and kidney function biomarkers.

```{r}
# Select variables for correlation analysis

corr_data <-
Diabetes %>%
select(
HbA1c,
Chol,
TG,
HDL,
LDL,
VLDL,
Urea,
Cr
)


# Calculate correlation matrix

cor_matrix <-
cor(corr_data,
    use = "complete.obs")


# Visualize correlations

corrplot(
cor_matrix,
method = "color",
type = "upper",
addCoef.col = "black",
tl.col = "black",
number.cex = 0.7
)
```

**Figure 3. Correlation matrix showing relationships between glycemic control, lipid profile, and kidney function biomarkers.**

---

## Relationship Between HbA1c and Individual Lipid Markers

To further examine specific relationships identified in the correlation analysis, scatterplots were generated between HbA1c and each lipid biomarker. Linear regression lines were added to visualize the direction of association.

---

## HbA1c and Total Cholesterol

```{r}
chol_plot <-

ggplot(Diabetes,
       aes(HbA1c,
           Chol)) +

geom_point() +

geom_smooth(
method = "lm",
se = TRUE
) +

labs(
title = "HbA1c vs Total Cholesterol",
x = "HbA1c (%)",
y = "Total Cholesterol"
) +

theme_classic()
```

---

## HbA1c and Triglycerides

```{r}
tg_plot <-

ggplot(Diabetes,
       aes(HbA1c,
           TG)) +

geom_point() +

geom_smooth(
method = "lm",
se = TRUE
) +

labs(
title = "HbA1c vs Triglycerides",
x = "HbA1c (%)",
y = "Triglycerides"
) +

theme_classic()
```

---

## HbA1c and HDL

```{r}
hdl_plot <-

ggplot(Diabetes,
       aes(HbA1c,
           HDL)) +

geom_point() +

geom_smooth(
method = "lm",
se = TRUE
) +

labs(
title = "HbA1c vs HDL",
x = "HbA1c (%)",
y = "HDL"
) +

theme_classic()
```

---

## HbA1c and LDL

```{r}
ldl_plot <-

ggplot(Diabetes,
       aes(HbA1c,
           LDL)) +

geom_point() +

geom_smooth(
method = "lm",
se = TRUE
) +

labs(
title = "HbA1c vs LDL",
x = "HbA1c (%)",
y = "LDL"
) +

theme_classic()
```

---

## HbA1c and VLDL

```{r}
vldl_plot <-

ggplot(Diabetes,
       aes(HbA1c,
           VLDL)) +

geom_point() +

geom_smooth(
method = "lm",
se = TRUE
) +

labs(
title = "HbA1c vs VLDL",
x = "HbA1c (%)",
y = "VLDL"
) +

theme_classic()
```

---

## Combined Lipid Profile Visualization

To improve visualization and allow direct comparison among lipid biomarkers, individual scatterplots were combined into a single multi-panel figure.

```{r}
ggarrange(
chol_plot,
tg_plot,
hdl_plot,
ldl_plot,
vldl_plot,
ncol = 2,
nrow = 3
)
```

**Figure 4. Association between HbA1c and lipid profile biomarkers.**

---

## Interpretation

The relationship between HbA1c and lipid biomarkers was evaluated based on correlation coefficients, scatterplot trends, and regression lines.

Possible interpretations include:

- A positive association between HbA1c and cholesterol, triglycerides, LDL, or VLDL would indicate that poorer glycemic control is associated with an unfavorable lipid profile.
- A negative association between HbA1c and HDL would suggest that worsening glycemic control is associated with reduced protective cholesterol levels.
- Weak or absent associations would suggest that additional factors may contribute to lipid variation among individuals with diabetes.


# Research Question 3

## Is HbA1c associated with kidney function?

Diabetic kidney disease is one of the most common complications of diabetes and is strongly influenced by prolonged exposure to elevated blood glucose levels. Chronic hyperglycemia can contribute to renal injury through oxidative stress, inflammation, and vascular dysfunction.

Serum urea and creatinine are commonly used biomarkers for evaluating kidney function. Therefore, this study examined whether HbA1c levels were associated with changes in kidney function markers.

---

## Relationship Between HbA1c and Kidney Function Biomarkers

Scatterplots with linear regression lines were generated to evaluate the association between HbA1c and kidney function markers.

---

## HbA1c and Urea

```{r}
urea_plot <-

ggplot(Diabetes,
       aes(HbA1c,
           Urea)) +

geom_point() +

geom_smooth(
method = "lm",
se = TRUE
) +

labs(
title = "HbA1c vs Urea",
x = "HbA1c (%)",
y = "Urea"
) +

theme_classic()

urea_plot
```

**Figure 5A. Relationship between HbA1c and serum urea levels.**

---

## HbA1c and Creatinine

```{r}
cr_plot <-

ggplot(Diabetes,
       aes(HbA1c,
           Cr)) +

geom_point() +

geom_smooth(
method = "lm",
se = TRUE
) +

labs(
title = "HbA1c vs Creatinine",
x = "HbA1c (%)",
y = "Creatinine"
) +

theme_classic()

cr_plot
```

**Figure 5B. Relationship between HbA1c and serum creatinine levels.**

---

## Combined Kidney Function Visualization

To improve comparison between renal biomarkers, both kidney function plots were displayed together.

```{r}
ggarrange(
urea_plot,
cr_plot,
ncol = 2
)
```

**Figure 5. Association between HbA1c and kidney function biomarkers.**

---

## Interpretation

The relationship between HbA1c and kidney function markers was evaluated based on the direction and strength of the regression trends.

A positive association between HbA1c and urea or creatinine would suggest that worsening glycemic control is associated with poorer kidney function. This would support the hypothesis that chronic hyperglycemia contributes to renal impairment.

However, weak or non-significant associations would suggest that kidney function is influenced by additional factors beyond glycemic control, including disease duration, age, medications, hypertension, and other metabolic factors.

# Research Question 4

## Can lipid profile and kidney function biomarkers predict HbA1c levels?

While individual correlations describe relationships between HbA1c and individual biomarkers, diabetes-related metabolic abnormalities often occur together. Therefore, a multiple linear regression model was performed to determine whether lipid profile markers and kidney function biomarkers collectively predict HbA1c levels.

The model included:

- Total cholesterol (Chol)
- Triglycerides (TG)
- HDL
- LDL
- VLDL
- Urea
- Creatinine (Cr)

HbA1c was used as the dependent variable because it represents long-term glycemic control.

---

## Multiple Linear Regression Model

The following model was used:

\[
HbA1c = \beta_0 + \beta_1(Chol) + \beta_2(TG) + \beta_3(HDL) + \beta_4(LDL) + \beta_5(VLDL) + \beta_6(Urea) + \beta_7(Cr)
\]

```{r}
model <-

lm(
HbA1c ~ Chol +
TG +
HDL +
LDL +
VLDL +
Urea +
Cr,

data = Diabetes
)


summary(model)
```
```{r}
# Extract regression coefficients as a table

summary(model)$coefficients


normalized_model <-

lm(
HbA1c ~ Chol +
TG +
HDL +
LDL +
VLDL +
Urea +
Cr,

data = Diabetes_normalized
)

summary(normalized_model)

```

---

## Regression Model Interpretation

The regression output provides several important measures:

### Model Fit

The **R-squared (R²)** value represents the proportion of variation in HbA1c explained by the combined lipid and kidney biomarkers.

A higher R² value indicates that the included biomarkers explain a greater proportion of differences in glycemic control among individuals.

---

### Predictor Contributions

The regression coefficients indicate the direction and magnitude of association between each biomarker and HbA1c after accounting for the influence of other variables.

Interpretation:

- Positive coefficients indicate that higher biomarker levels are associated with higher HbA1c.
- Negative coefficients indicate an inverse relationship with HbA1c.
- Significant predictors (p < 0.05) represent biomarkers independently associated with glycemic control after adjustment for other variables.

---

## Summary of Predictive Analysis

This analysis evaluates whether abnormalities in lipid metabolism and kidney function collectively contribute to differences in glycemic control.

If significant predictors are identified, these findings would suggest that metabolic and renal biomarkers may provide additional information regarding diabetes severity beyond HbA1c alone.

However, because this analysis is based on observational data, associations should not be interpreted as causal relationships.



```


## References

Add any references to papers, data sources, or resources you used.

------------------------------------------------------------------------

**Last updated:** June 2026
